---
title: "Raster and Vector Data Cubes Across Spatial Data Science Languages"
subtitle: "Introduction to vector data cubes in Python, R and Julia"
date: 2025-06-23
author: 
  - name: Yomna Eid
    affiliations:
    - ref: unimün
  - name: Mohammad Alasawedah
    affiliations: 
    - ref: unimün
  - name: Abhishek Singh
    affiliations: 
    - ref: eurac
  - name: Felix Cremer
    affiliations: 
    - ref: mpi
  - name: Lazaro Alonso
    affiliations: 
    - ref: mpi
affiliations:
- id: unimün
  name: University of Münster 
- id: eurac
  name: Eurac Research
- id: mpi
  name: Max-Planck Institute for Biogeochemistry
format:
  html:
    code-fold: false
  pdf:
    pdf-engine: xelatex
theme:
  light: flatly
  dark: darkly
toc: true
number-sections: true
documentclass: article
engine: julia
julia:
  exeflags: ["--project=@."]
---

# Introdution to Raster and Vector Data Cubes

In this notebook, we convert a raster data cube covering a large spatial extent to a vector data cube containing only the spatial areas of interest. Observe how geospatial features, or data associated with those features, change over time. 

Here we present the code for analysis in R and Python.

## Install PythonCall and RCall

When using the `julia` engine use the above packages for a better interoperability across languages.

::: {.panel-tabset}

## Installation
```{julia}
#| output: false
#| eval: true
using Pkg
Pkg.add("PythonCall")
Pkg.add("RCall")
Pkg.add("CondaPkg")
```

## Imports
```{julia}
#| output: false
import PythonCall
import RCall
```
:::


## Install and loading packages for Julia, R and Python

::: {.panel-tabset group="language"}

## Julia

```{julia}
#| output: true
#| eval: true
Pkg.add([
    "Glob",
    "Dates",
    "DataFrames",
    "DimensionalData",
    "Shapefile",
    "GeometryOps",
    "ArchGDAL",
    "GeoInterface",
    "Proj",
    "Statistics",
    "CairoMakie",
    "LibGEOS",
    "IntervalSets",
    "Tyler",
    ]
    )

#the bleeding edge devs (soon to be merge)
Pkg.add(url="https://github.com/rafaqz/Rasters.jl", rev="as/vectordatacubes")
```

start using the packages

```{julia}
using Glob
using Dates
using Statistics
using DimensionalData
using DataFrames
# rasters
using Rasters
using Shapefile
using Proj
import ArchGDAL
import GeometryOps as GO
using IntervalSets
import GeoInterface as GI
import LibGEOS
# for plotting
using CairoMakie
using Tyler
using Tyler.TileProviders  # for map tiles providers
set_theme!(theme_light(); size = (650, 350), fontsize=12.5)
```

## R

Note the following:

- On MacOS you first need to install https://www.xquartz.org !
- The following has been tested with `R 4.5.1 (2025-06-13)`.


```{r}
# If the libraries are not available, install them
packages <- c("terra", "lubridate", "stars", "ggplot2", "caret", "ranger", "dplyr", "tidyr")
for (p in packages){
  if (!requireNamespace(p, quietly = TRUE)){
    install.packages(p, repos = "https://cran.uni-muenster.de/")
  }
}
lapply(packages, library, character.only = TRUE);
```

## Python

We install Python packages via `CondaPkg.jl`
```{julia}
#| output: true
#| eval: true
using CondaPkg
CondaPkg.add([
    "numpy",
    "pandas",
    "xarray",
    "geopandas",
    "rasterio",
    "rioxarray",
    "shapely",
    "matplotlib",
    "folium",
    "leafmap",
    ]
    )
```

Now, support for Vector data cubes in Xarray is provided by `Xvec`, which can also be installed via the `conda-forge` channel. See https://xvec.readthedocs.io/en/stable/#installing for more information.

```{julia}
#| output: false
#| eval: true
CondaPkg.add("xvec")
```

And import them as usual

```{python}
import os
import string
from glob import glob
import pandas as pd
import numpy as np
import xarray as xr
import geopandas as gpd
import xvec
# import rasterio
# import rioxarray as rio
import shapely.geometry
from shapely.geometry import Point
import matplotlib.pyplot as plt
import matplotlib.cm as cm
import folium
import leafmap.foliumap as leafmap
```

:::

Now, let's test the 3 languages!

::: {.panel-tabset group="language"}

## Julia
Create some data in Julia:

```{julia}
data = sin.(0:0.1:4pi);
```

## R
Then plot it in R by interpolating the `data` variable from Julia into `R` via
the `$` syntax:

```{r}
plot($data)
```

The same `$` syntax can be used to interpolate Julia variables into Python code as well:

## Python
```{python}
len($data)
```

```{python}
import numpy as np
np.sqrt(len($data))
```

:::

## Load files

```{julia}
path_lai = joinpath(@__DIR__, "../Data/lai_cropped/");
```

```{julia}
#define functions to read the data + read dates from the file name
function read_files(path)
    ras_files = glob("*.tif", path)
    ras_dates = get_dates.(ras_files)
    ras = Raster.(ras_files, lazy=true, name=:data)
    return cat(ras..., dims=Ti(ras_dates))
end
function get_dates(x)
    return DateTime(split(x, r"[_.]")[end-1], "yyyymmddTHHMMSS")
end

lai_ds = read_files(path_lai)
```

## Plot one time stamp

```{julia}
plot(lai_ds[Ti=1]; colormap=:magma);
```

## Plots

And now

```{julia}
# london
provider = Tyler.TileProviders.Esri(:WorldImagery)
london = Rect2f(-0.0921, 51.5, 0.04, 0.025)

with_theme(theme_dark()) do
  m=Tyler.Map(london; provider, figure=Figure(; size=(650,400)))
  m.axis.aspect=650/400
  hidedecorations!(m.axis)
  hidespines!(m.axis)
  m
end
```
